True versus apparent shapes of bow shocks

### 2018-01-18 papers/Papers1234/sec-projection.tex
Explain about the plane of sky as a projective space

Consider body-frame cartesian coordinates $(x,y,z)$, where \(x\) is
the symmetry axis, and spherical polar coordinates
\((R, \theta, \phi)\), where \(\theta\) is the polar angle and
\(\phi\) the azimuthal angle.  Since the surface is cylindrically
symmetric, it is can be specified as $R = R(\theta)$, so that
cartesian coordinates on the surface are:
\begin{equation}
  \bm{R} \equiv
  \begin{Vector}
    x \\ y \\ z
  \end{Vector} 
  = 
  \begin{Vector}
    R(\theta)\, \cos\theta \\
    R(\theta)\, \sin\theta\cos\phi \\
    R(\theta)\, \sin\theta\sin\phi
  \end{Vector}.
  \label{eq:body-frame}
\end{equation} 
Suppose that the viewing direction makes an angle \(i\) with the
\(z\)~axis, so that we can define observer-frame coordinates
\((x', y', z')\), which are found by rotating the body-frame
coordinates about the \(y\)~axis:
\begin{equation}
  \bm{R}' \equiv
  \begin{Vector}
    x' \\ y' \\ z'
  \end{Vector}
  = \mathbfss{A}_y(i)\, \bm{R}
  =
  \begin{Vector}
    x\cos i - z\sin i\\
    y \\
    z\cos i + x\sin i
  \end{Vector} \ ,
  \label{eq:Trans}
\end{equation}
where the rotation matrix \(\mathbfss{A}_y(i)\) is given in Appendix~\ref{sec:plane-sky-projection}. 
The relationship between the two frames is illustrated in
Figure~\ref{fig:projection-pos}.  All quantities in the observer's
frame are denoted by attaching a prime to the equivalent quantity in
the body frame.  The celestial coordinates on the ``plane of the sky''
are given by \((x', y')\) with \(x'\) being the projected symmetry
axis of the surface, while the line of sight lies along \(-z'\).  The
inclination angle \(i\) is defined so that \(i = \ang{0}\) when the
surface is viewed perpendicular to its axis (\textit{side on}) and
\(i = \ang{90}\) when it is viewed along its axis (\textit{end on}).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/projection-pos}
  \caption{Relationship between body frame (unprimed coordinates) and
    observer frame (primed coordinates).}
  \label{fig:projection-pos}
\end{figure}

@@

The boundary on the plane of the sky of the projected surface is the
locus of those lines of sight that graze the surface tangentially.
This corresponds to a curved line on the surface itself, which we
denote the \textit{tangent line}, and which is defined by the
condition  
\begin{equation}
  \label{eq:tangent-line-condition}
  \uvec{n}'\cdot \uvec{z}' = 0.
\end{equation}
We denote by \(\phi\T\) that value of $\phi$ that satisfies this
relation for a given inclination, \(i\), and polar angle, \(\theta\):
\begin{equation}
\sin\phi\T = -\tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}
From equations~(\ref{eq:body-frame}, \ref{eq:Trans}) it follows that
the observer-frame coordinates of the tangent line are given by
\begin{equation}
\left(\begin{array}{c}
x'\T \\ y'\T \\ z'\T
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi\T \sin i \\
\sin\theta(1-\sin^2\phi\T)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi\T\cos i
\end{array}\right).
\label{eq:tangential}
\end{equation} 
Note that, in general, \(z'\T\) is not a linear function of \(x'\T\)
and \(y'\T\), so that the tangent line is not a plane curve.

The apparent shape \((x'\T, y'\T)\) of the tangent line on the plane
of the sky can also be described in polar form as \(R'(\theta')\),
where
\begin{equation}
  \label{eq:R-prime-theta-prime}
  R' = (x'^2\T+ y'^2\T)^{1/2} 
  \quad \text{and} \quad
  \tan\theta' = y'\T / x'\T.
\end{equation}
Note that equation~(\ref{eq:tanphi}) will not have
a solution for arbitrary values of $\theta$ and $i$, but only when
$|\tan i\tan\alpha|<1$. In particular, if $i\neq 0$, then the tangent
line only exists for \(\theta > \theta_{0}\) where \(\theta_{0}\) is the value of \(\theta\) on the tangent line's projected symmetry axis (\(\theta' = 0\)), 
given implicitly by
\begin{align}
\tan\theta_{0} = \frac{|\tan i| + \omega(\theta_{0})}{1-\omega(\theta_{0}) |\tan i|} . 
\label{eq:thetapar}
\end{align}
In addition, if the surface is sufficiently ``open''
(\(\alpha \ge \alpha_{\mathrm{min}} > 0\) for all \(\theta\)), then for those
inclinations with
\(\vert i\vert > (\ang{90} - \alpha_{\mathrm{min}}) \) the tangent line does not exist
for any value of \(\theta\).  In other words, when the viewing angle is
sufficiently close to face-on, the projected surface has no ``edge''
and will no longer look like a bow shock to the observer.

@@

>>>>

Consider body-frame cartesian coordinates $(x,y,z)$, where \(x\) is
the symmetry axis, and spherical polar coordinates
\((R, \theta, \phi)\), where \(\theta\) is the polar angle and
\(\phi\) the azimuthal angle.  Since the surface is cylindrically
symmetric, it is can be specified as $R = R(\theta)$, so that
cartesian coordinates on the surface are:
\begin{equation}
  \bm{r} \equiv
  \begin{Vector}
    x \\ y \\ z
  \end{Vector} 
  = 
  \begin{Vector}
    R(\theta)\, \cos\theta \\
    R(\theta)\, \sin\theta\cos\phi \\
    R(\theta)\, \sin\theta\sin\phi
  \end{Vector}.
  \label{eq:body-frame}
\end{equation} 
Suppose that the viewing direction makes an angle \(i\) with the
\(z\)~axis, so that we can define observer-frame coordinates
\((x', y', z')\), which are found by rotating the body-frame
coordinates about the \(y\)~axis:
\begin{equation}
  \bm{r}' \equiv
  \begin{Vector}
    x' \\ y' \\ z'
  \end{Vector}
  = \mathbfss{A}_y(i)\, \bm{r}
  =
  \begin{Vector}
    x\cos i - z\sin i\\
    y \\
    z\cos i + x\sin i
  \end{Vector} \ ,
  \label{eq:Trans}
\end{equation}
where the rotation matrix \(\mathbfss{A}_y(i)\) is given in
Appendix~\ref{sec:plane-sky-projection}.  The inclination angle \(i\)
is defined so that \(i = \ang{0}\) when the surface is viewed
perpendicular to its axis (\textit{side on}) and \(i = \ang{90}\) when
it is viewed along its axis (\textit{end on}).

The relationship between the two frames is illustrated in
Figure~\ref{fig:projection-pos}.  All quantities in the observer's
frame are denoted by attaching a prime to the equivalent quantity in
the body frame. There are two ways of interpreting the primed
coordinates. On the one hand, the 3-vector \((x', y', z')\) specifies
a point in Euclidean space, \(\mathds{R}^3\), but an alternative
interpretation is to take the 2-vector \((x', y')\) as specifying a
point in a \textit{projective space}, \(\mathds{P}^2\) (see, for
example, \S~15.6 of \citealp{Penrose:2004a}). Each ``point'' in
\(\mathds{P}^2\) is equivalent to a line in \(\mathds{R}^3\),
specifically: a line of sight that passes through the observer. Thus,
\((x', y')\) gives the celestial coordinates on the \textit{plane of
  the sky}, with \(x'\) being the projected symmetry axis of the
surface.  We assume that the observer is located at a very large
distance, relative to the size of the bow, so that all lines of sight
are effectively parallel to the \(z'\) axis, with the observer at
\(z' = -\infty\).  But, from the point of view of the plane of the sky, the
\(z'\) coordinate is strictly irrelevant since it is a projective
plane, and not a Euclidean plane.  In the following, we will switch
between the \(\mathds{R}^3\) and \(\mathds{P}^2\) interpretations as
convenient, resolving ambiguity where necessary via the adjectives
``Euclidean'' for \(\mathds{R}^3\) and ``plane-of-sky'' or
``projected'' for \(\mathds{P}^2\).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{figs/projection-pos}
  \caption{Relationship between body frame (unprimed coordinates) and
    observer frame (primed coordinates).  Note that the plane of the
    sky is a projective plane, not a geometric plane in Euclidean
    3-space, see discussion in text.}
  \label{fig:projection-pos}
\end{figure}

@@

The boundary on the plane of the sky of the projected surface is the
locus of those lines of sight that graze the surface tangentially.
This corresponds to a curved line on the surface itself, which we
denote the \textit{tangent line}, and which is defined by the
condition  
\begin{equation}
  \label{eq:tangent-line-condition}
  \uvec{n}'\cdot \uvec{z}' = 0.
\end{equation}
We denote by \(\phi\T\) that value of $\phi$ that satisfies this
relation for a given inclination, \(i\), and polar angle, \(\theta\):
\begin{equation}
\sin\phi\T = -\tan i\tan\alpha = \tan i \frac{1+\omega\tan\theta}{\omega-\tan\theta}
\label{eq:tanphi}
\end{equation}
From equations~(\ref{eq:body-frame}, \ref{eq:Trans}) it follows that
the observer-frame coordinates of the tangent line are given by
\begin{equation}
\left(\begin{array}{c}
x'\T \\ y'\T \\ z'\T
\end{array}\right)= R(\theta)\left(\begin{array}{c}
\cos\theta\cos i - \sin\theta\sin\phi\T \sin i \\
\sin\theta(1-\sin^2\phi\T)^{1/2} \\
\cos\theta\sin i +\sin\theta\sin\phi\T\cos i
\end{array}\right).
\label{eq:tangential}
\end{equation} 
Note that, in general, \(z'\T\) is not a linear function of \(x'\T\)
and \(y'\T\), so that the tangent line \((x'\T, y'\T, z'\T)\) is not a
plane curve in 3-dimensional Euclidean space, \(\mathds{R}^3\).
However, for the projected shape \((x'\T, y'\T)\) of the tangent line
on the plane of the sky, \(\mathds{P}^2\), the value of \(z'\T\) does
not matter (see above).  The projected shape can also be described in
polar form as \(R'(\theta')\), where
\begin{equation}
  \label{eq:R-prime-theta-prime}
  R' = (x'^2\T+ y'^2\T)^{1/2} 
  \quad \text{and} \quad
  \tan\theta' = y'\T / x'\T.
\end{equation}

Note that equation~(\ref{eq:tanphi}) will not have
a solution for arbitrary values of $\theta$ and $i$, but only when
$|\tan i\tan\alpha|<1$. In particular, if $i\neq 0$, then the tangent
line only exists for \(\theta > \theta_{0}\) where \(\theta_{0}\) is the value of \(\theta\) on the tangent line's projected symmetry axis (\(\theta' = 0\)), 
given implicitly by
\begin{align}
\tan\theta_{0} = \frac{|\tan i| + \omega(\theta_{0})}{1-\omega(\theta_{0}) |\tan i|} . 
\label{eq:thetapar}
\end{align}
In addition, if the surface is sufficiently ``open''
(\(\alpha \ge \alpha_{\mathrm{min}} > 0\) for all \(\theta\)), then for those
inclinations with
\(\vert i\vert > (\ang{90} - \alpha_{\mathrm{min}}) \) the tangent line does not exist
for any value of \(\theta\).  In other words, when the viewing angle is
sufficiently close to face-on, the projected surface has no ``edge''
and will no longer look like a bow shock to the observer.

@@

Projective space \(\mathbb{RP}^2\) or \(\mathbbm{RP}^2\) or \(\mathds{RP}^2\)
